---
title: "2 AT8 - 01 Prep"
author:
  - name: "Emir Turkes"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
link-citations: true
output:
  html_document:
    code_folding: hide
    number_sections: true
    theme: lumen
    highlight: haddock
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
knit: (function(inputFile, encoding) {
  rmarkdown::render(
    inputFile, encoding = encoding,
    output_file = file.path(
      "..", "..", "results", "individual", "2_AT8_01prep.html"
    ))})
---

<style type="text/css">
body {font-size: 16px;}
h1.title {font-size: 35px;}
h1 {font-size: 24px;}
h2 {font-size: 22px;}
.toc-content {padding-left: 0px; padding-right: 0px;}
div.tocify {width: 100%;}
.tocify-subheader .tocify-item {
  font-size: 0.95em; padding-left: 25px; text-indent: 0;
}
.tocify-subheader .tocify-subheader .tocify-item {
  font-size: 0.95em; padding-left: 35px; text-indent: 0;
}
div.main-container {max-width: none; width: 100%;}
</style>

*This file is a part of [otero-garcia-2022-ssRNAseq](https://github.com/eturkes/otero-garcia-2022-ssRNAseq).*

The table of contents in the top left is clickable and can be used to quickly navigate the document.
To toggle the visibility of code, use the `CODE` toggles at the top right of chunks.
The toggle at the start of the document controls the visibility of all chunks.

```{r}
#    This file is part of otero-garcia-2022-ssRNAseq.
#    Copyright (C) 2024  Emir Turkes, UK DRI at UCL
#
#    This program is free software: you can redistribute it and/or modify
#    it under the terms of the GNU General Public License as published by
#    the Free Software Foundation, either version 3 of the License, or
#    (at your option) any later version.
#
#    This program is distributed in the hope that it will be useful,
#    but WITHOUT ANY WARRANTY; without even the implied warranty of
#    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
#    GNU General Public License for more details.
#
#    You should have received a copy of the GNU General Public License
#    along with this program.  If not, see <http://www.gnu.org/licenses/>.
#
#    Emir Turkes can be contacted at emir.turkes@eturkes.com

# Load required packages, suppressing startup messages.
# -----------------------------------------------------
library(conflicted)
packages <- c(
  "Seurat", "parallelly", "ggplot2", "dplyr", "ggrepel", "viridis",
  "DropletUtils", "scDblFinder", "BiocParallel", "DT", "scuttle"
)
invisible(
  suppressPackageStartupMessages(
    lapply(packages, FUN = library, character.only = TRUE)
  )
)
# -----------------------------------------------------

# Define global settings.
# -----------------------
knitr::opts_chunk$set(fig.width = 10, fig.height = 7, dpi = 96)
# -----------------------

# Define functions.
# -----------------
source(file.path("..", "utils.R"))
# -----------------

# Useful variables.
# -----------------
name <- "2_AT8"
step <- 1

data_dir <- file.path("..", "..", "data", "raw_data")
cache_dir <- file.path("..", "..", "cache", name, paste0("step", "0", step))
if (!dir.exists(cache_dir)) {
  dir.create(cache_dir, recursive = TRUE)
}

protocol <- c("human", "droplet", "single-cell", "umis")
vars_to_regress <- NULL
parallel_override <- NULL
# -----------------

# Metadata to plot after dimensionality reduction and clustering.
# Values in list can include "no_legend and/or "no_label" to exclude those.
# -------------------------------------------------------------------------
metadata_to_plot <- vector("list", length = 2)
names(metadata_to_plot) <- c("seurat_clusters", "Phase")
metadata_to_plot[[2]] <- "no_label"
# -------------------------------------------------------------------------
```

# Cleaning

We first read in the data and form a Seurat object.

```{r}
seurat <- CreateSeuratObject(
  Read10X_h5(
    file.path(data_dir, "GSM3704360_2-AT8_filtered_feature_bc_matrix.h5"),
    use.names = FALSE
  )
)

seurat@misc <- list(
  geneids = rownames(
    Read10X_h5(
      file.path(data_dir, "GSM3704360_2-AT8_filtered_feature_bc_matrix.h5")
    )
  )
)
gene_anno <- data.frame(
  symbol = unlist(seurat@misc), ensembl = rownames(seurat)
)

seurat
```

# QC

In a droplet-based experiment, we first remove empty droplets.
Since they are removed already, we just visualise it here.

```{r}
if (protocol[2] == "droplet") {
  bcrank <- barcodeRanks(GetAssayData(seurat), lower = 20)
  uniq <- !duplicated(bcrank$rank)
  plot(
    bcrank$rank[uniq], bcrank$total[uniq], log = "xy",
    xlab = "Rank", ylab = "Total UMI count", cex.lab = 1.2
  )
  abline(h = metadata(bcrank)$inflection, col = "darkgreen", lty = 2)
  abline(h = metadata(bcrank)$knee, col = "dodgerblue", lty = 2)
  legend(
    "bottomleft", legend = c("Inflection", "Knee"),
    col = c("darkgreen", "dodgerblue"),
    lty = 2, cex = 1.2
  )
  rm(bcrank)
}
```

We remove cells low in unique features and total counts and calculate percentage of mitochondrial and ribosomal reads and other genes commonly used as indicators of low-quality cells.
We also use mitochondrial reads as a proxy for ambient RNA contamination.

```{r}
mito <- grep("^MT-", gene_anno$symbol)
datatable_download(gene_anno[mito, ])

is_mito <- rownames(seurat) %in% gene_anno[mito, ]$ensembl
mito <- which(rownames(seurat) %in% gene_anno[mito, ]$ensembl)
seurat[["percent_mt"]] <- PercentageFeatureSet(
  seurat, features = rownames(seurat)[mito]
)

ribo <- grep("^RP[SL][[:digit:]]", gene_anno$symbol)
datatable_download(gene_anno[ribo, ])

ribo <- which(rownames(seurat) %in% gene_anno[ribo, ]$ensembl)
seurat[["percent_ribo"]] <- PercentageFeatureSet(
  seurat, features = rownames(seurat)[ribo]
)

seurat[["percent_malat1"]] <- PercentageFeatureSet(
  seurat, features = "ENSG00000251562"
)

if (length(mito) != 0) {
  if (protocol[3] == "single-nuc") {
    print(paste0("Removed ", length(mito), " mitochondrial genes:"))
    seurat <- seurat[-mito, ]
  } else {
    print(
      paste0("Found but did not remove ", length(mito), " mitochondrial genes:")
    )
  }
}

VlnPlot(
  seurat,
  features = c(
    "nFeature_RNA", "nCount_RNA", "percent_mt", "percent_ribo", "percent_malat1"
  ),
  ncol = 5
)

seurat@meta.data %>%
  ggplot(aes(x = nFeature_RNA)) +
  geom_density(alpha = 0.2) +
  scale_x_log10() +
  theme_classic() +
  ylab("Cell Density")

seurat@meta.data %>%
  ggplot(aes(x = nCount_RNA)) +
  geom_density(alpha = 0.2) +
  scale_x_log10() +
  theme_classic() +
  ylab("Cell Density")

seurat@meta.data %>%
  ggplot(aes(x = percent_mt)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  ylab("Cell Density")

seurat@meta.data %>%
  ggplot(aes(x = percent_ribo)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  ylab("Cell Density")

seurat@meta.data %>%
  ggplot(aes(x = percent_malat1)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  ylab("Cell Density")
```

We perform doublet detection in a droplet-based experiments.

```{r}
if (protocol[2] == "droplet") {

  rds <- file.path(cache_dir, "doublets.rds")
  if (file.exists(rds)) {
    sub_name <- "doublets"
    seurat <- cluster_pipeline(
      seurat, cache_dir = cache_dir,
      sub_name = sub_name, protocol = protocol,
      vars_to_regress = vars_to_regress, parallel_override = parallel_override,
      res_divider = 20000
    )
  } else {
    sce <- SingleCellExperiment(list(counts = GetAssayData(seurat)))
    sce <- scDblFinder(sce, BPPARAM = MulticoreParam(availableCores()))
    doublets <- sce$scDblFinder.class
    rm(sce)
    seurat$doublets <- ifelse(doublets == "doublet", yes = 1, no = 0)

    sub_name <- "doublets"
    seurat <- cluster_pipeline(
      seurat, cache_dir = cache_dir,
      sub_name = sub_name, protocol = protocol,
      vars_to_regress = vars_to_regress, parallel_override = parallel_override,
      res_divider = 20000
    )
  }

  print(seurat)
  FeaturePlot(seurat, features = "doublets")
}
```

# Cluster Pipeline

We run a pipeline that applies normalisation and scaling, dimensionality reduction, and clustering.
Before running it, doublets and cells with 10% or greater percentage of mitochondrial reads are dropped.
We also remove cells that are outliers in more than 3 MADs in terms of library size and number of unique features.

```{r}
remove <- which(seurat$doublets == 1)
print(paste0("Remove doublets: ", length(remove)))
seurat <- seurat[ , -remove]
seurat

seurat <- subset(seurat, subset = percent_mt < 10)
print("Remove cells with 10% and over percent mitochondrial reads")
seurat

sce <- SingleCellExperiment(
  list(counts = GetAssayData(seurat, assay = "RNA", layer = "counts"))
)
remove <- perCellQCFilters(perCellQCMetrics(sce))
rm(sce)
seurat <- seurat[ , !remove$discard]
seurat

sub_name <- "prep"
seurat <- cluster_pipeline(
  seurat, cache_dir = cache_dir,
  sub_name = sub_name, protocol = protocol,
  vars_to_regress = vars_to_regress, parallel_override = parallel_override,
  cc = FALSE, res_divider = 50000, min_cells = 0
)
seurat

for (i in seq_along(metadata_to_plot)) {
  if (
    "no_legend" %in% metadata_to_plot[[i]] &&
      "no_label" %in% metadata_to_plot[[i]]
  ) {
    print(
      red_dim_plot(
        seurat, x = "umap1", y = "umap2", color = names(metadata_to_plot)[i]
      ) +
        NoLegend() +
        ggtitle(names(metadata_to_plot)[i])
    )
  } else if ("no_legend" %in% metadata_to_plot[[i]]) {
    print(
      red_dim_plot(
        seurat, x = "umap1", y = "umap2",
        color = names(metadata_to_plot)[i], type = "cat"
      ) +
        NoLegend() +
        ggtitle(names(metadata_to_plot)[i])
    )
  } else if ("no_label" %in% metadata_to_plot[[i]]) {
    print(
      red_dim_plot(
        seurat, x = "umap1", y = "umap2", color = names(metadata_to_plot)[i]
      ) +
        ggtitle(names(metadata_to_plot)[i])
    )
  } else {
    print(
      red_dim_plot(
        seurat, x = "umap1", y = "umap2",
        color = names(metadata_to_plot)[i], type = "cat"
      ) +
        ggtitle(names(metadata_to_plot)[i])
    )
  }
}

red_dim_plot(
  seurat, x = "umap1", y = "umap2", color = "nFeature_SCT", type = "cont"
) + ggtitle("nFeature_SCT")
red_dim_plot(
  seurat, x = "umap1", y = "umap2", color = "nCount_SCT", type = "cont"
) + ggtitle("nCount_SCT")
red_dim_plot(
  seurat, x = "umap1", y = "umap2", color = "nFeature_RNA", type = "cont"
) + ggtitle("nFeature_RNA")
red_dim_plot(
  seurat, x = "umap1", y = "umap2", color = "nCount_RNA", type = "cont"
) + ggtitle("nCount_RNA")

red_dim_plot(
  seurat, x = "umap1", y = "umap2", color = "percent_mt", type = "cont"
) + ggtitle("percent_mt")
red_dim_plot(
  seurat, x = "umap1", y = "umap2", color = "percent_ribo", type = "cont"
) + ggtitle("percent_ribo")
red_dim_plot(
  seurat, x = "umap1", y = "umap2", color = "percent_ribo", type = "cont"
) + ggtitle("percent_malat1")

VlnPlot(
  seurat,
  features = c("nFeature_RNA", "nCount_RNA"),
  ncol = 2
)
VlnPlot(
  seurat,
  features = c("nFeature_SCT", "nCount_SCT"),
  ncol = 2
)
VlnPlot(
  seurat,
  features = c("percent_mt", "percent_ribo"),
  ncol = 2
)
```

# References

This is the concluding section of the document, where we output the `sessionInfo`, and create a bibliography for works cited.

```{r}
sessionInfo()
```
