---
title: "Otero-Garcia 2022 ssRNAseq - 03 Testing"
author:
  - name: "Emir Turkes"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
link-citations: true
output:
  html_document:
    code_folding: hide
    number_sections: true
    theme: lumen
    highlight: haddock
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
knit: (function(inputFile, encoding) {
  rmarkdown::render(
    inputFile, encoding = encoding,
    output_file = file.path(
      "..", "..", "results",
      "comb", "otero_garcia_2022_ssRNAseq_03testing.html"
    ))})
---

<style type="text/css">
body {font-size: 16px;}
h1.title {font-size: 35px;}
h1 {font-size: 24px;}
h2 {font-size: 22px;}
.toc-content {padding-left: 0px; padding-right: 0px;}
div.tocify {width: 100%;}
.tocify-subheader .tocify-item {
  font-size: 0.95em; padding-left: 25px; text-indent: 0;
}
.tocify-subheader .tocify-subheader .tocify-item {
  font-size: 0.95em; padding-left: 35px; text-indent: 0;
}
div.main-container {max-width: none; width: 100%;}
</style>

*This file is a part of [otero-garcia-2022-ssRNAseq](https://github.com/eturkes/otero-garcia-2022-ssRNAseq).*

The table of contents in the top left is clickable and can be used to quickly navigate the document.
To toggle the visibility of code, use the `CODE` toggles at the top right of chunks.
The toggle at the start of the document controls the visibility of all chunks.

```{r}
#    This file is part of otero-garcia-2022-ssRNAseq.
#    Copyright (C) 2024  Emir Turkes, UK DRI at UCL
#
#    This program is free software: you can redistribute it and/or modify
#    it under the terms of the GNU General Public License as published by
#    the Free Software Foundation, either version 3 of the License, or
#    (at your option) any later version.
#
#    This program is distributed in the hope that it will be useful,
#    but WITHOUT ANY WARRANTY; without even the implied warranty of
#    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
#    GNU General Public License for more details.
#
#    You should have received a copy of the GNU General Public License
#    along with this program.  If not, see <http://www.gnu.org/licenses/>.
#
#    Emir Turkes can be contacted at emir.turkes@eturkes.com

# Load required packages, suppressing startup messages.
# -----------------------------------------------------
library(conflicted)
conflicts_prefer(edgeR::cpm, .quiet = TRUE)
packages <- c("GSEABase", "Seurat", "scuttle", "edgeR", "IHW")
invisible(
  suppressPackageStartupMessages(
    lapply(packages, FUN = library, character.only = TRUE)
  )
)
# -----------------------------------------------------

# Define global settings.
# -----------------------
knitr::opts_chunk$set(fig.width = 10, fig.height = 7, dpi = 96)
# -----------------------

# Useful variables.
# -----------------
name <- "comb"
step <- 3

cache_dir <- file.path("..", "..", "cache", name, paste0("step", "0", step))
if (!dir.exists(cache_dir)) {
  dir.create(cache_dir, recursive = TRUE)
}
# -----------------
```

# Prep

```{r}
seurat <- readRDS(file.path(cache_dir, "..", "step02", "final_seurat.rds"))
seurat

ctypes <- list("EX L2-4 CUX2", "EX L4-5 RORB", "EX L5 RORB", "EX L5-6 THEMIS")
seurat <- subset(seurat, idents = ctypes)
gc()
seurat

gene_sets <- readRDS(
  file.path(cache_dir, "..", "step02", "gene_sets.rds")
)
gene_sets_keep <- readRDS(
  file.path(cache_dir, "..", "step02", "gene_sets_keep.rds")
)
gene_anno <- readRDS(file.path(cache_dir, "..", "step02", "gene_anno.rds"))
```

# Testing

```{r}
gse_data <- vector("list", length = 4)
deg_data <- vector("list", length = 4)
gse_list <- vector("list", length = 4)
deg_list <- vector("list", length = 4)

for (i in seq_along(ctypes)) {

  gse <- as.SingleCellExperiment(
    subset(seurat, idents = ctypes[i]), assay = "GF"
  )
  gse <- gse[rownames(gse) %in% names(gene_sets_keep), ]
  gse <- suppressWarnings(
    aggregateAcrossCells(
      gse, colData(gse)[ , "donor_tau"],
      use_exprs_values = "counts", statistics = "sum"
    )
  )
  gse <- gse[ , order(gse$tau)]

  deg <- as.SingleCellExperiment(
    subset(seurat, idents = ctypes[i]), assay = "RNA"
  )
  deg <- suppressWarnings(
    aggregateAcrossCells(
      deg, colData(deg)[ , "donor_tau"],
      use_exprs_values = "counts", statistics = "sum"
    )
  )
  deg <- deg[ , order(deg$tau)]

  dge_orig <- DGEList(counts(gse))
  keep <- filterByExpr(
    dge_orig, group = gse$tau,
    min.count = 100, min.total.count = 150
  )
  dge_orig <- dge_orig[keep, ]
  dge_orig <- calcNormFactors(dge_orig)
  dge <- DGEList(counts(gse))
  dge$samples$norm.factors <- dge_orig$samples$norm.factors
  keep <- filterByExpr(
    dge, group = gse$tau,
    min.count = 100, min.total.count = 150
  )
  dge <- dge[keep, , keep.lib.sizes = FALSE]
  dge <- calcNormFactors(dge)

  gse <- gse[rownames(gse) %in% rownames(dge)]
  logcounts(gse, withDimnames = FALSE) <- cpm(
    dge, log = TRUE, prior.count = 0.3
  )
  gse_data[[i]] <- gse

  gene_sets_sub <- gene_sets[names(gene_sets) %in% rownames(gse)]
  gene_sets_sub <- unique(unlist(geneIds(gene_sets_sub)))
  deg <- deg[rownames(deg) %in% gene_sets_sub, ]
  dge_orig <- DGEList(counts(deg))
  keep <- filterByExpr(
    dge_orig, group = deg$tau,
    min.count = 50, min.total.count = 75
  )
  dge_orig <- dge_orig[keep, ]
  dge_orig <- calcNormFactors(dge_orig)
  dge <- DGEList(counts(deg))
  dge$samples$norm.factors <- dge_orig$samples$norm.factors
  keep <- filterByExpr(
    dge, group = deg$tau,
    min.count = 50, min.total.count = 75
  )
  dge <- dge[keep, , keep.lib.sizes = FALSE]
  dge <- calcNormFactors(dge)

  deg <- deg[rownames(deg) %in% rownames(dge)]
  logcounts(deg, withDimnames = FALSE) <- cpm(
    dge, log = TRUE, prior.count = 0.1
  )
  deg_data[[i]] <- deg

  design <- model.matrix(~ 0 + gse$tau)
  colnames(design) <- make.names(colnames(design))
  cont_mat <- makeContrasts(
    NFT_CTRL =
      gse.tauNFT -
      gse.tauCTRL,
    levels = design
  )

  corr <- duplicateCorrelation(logcounts(gse), design, block = gse$donor)
  fit <- lmFit(
    logcounts(gse), design,
    correlation = corr$consensus.correlation, block = gse$donor
  )
  fit <- contrasts.fit(fit, cont_mat)
  fit <- eBayes(fit, trend = TRUE, robust = TRUE)
  print(plotSA(fit))

  tests <- decideTests(fit, method = "global")
  write.fit(
    fit, tests, file.path(cache_dir, "results.tsv"),
    adjust = "BH", F.adjust = "BH", method = "global"
  )
  gse_list[[i]] <- read.delim(file.path(cache_dir, "results.tsv"))

  corr <- duplicateCorrelation(logcounts(deg), design, block = deg$donor)
  fit <- lmFit(
    logcounts(deg), design,
    correlation = corr$consensus.correlation, block = deg$donor
  )
  fit <- contrasts.fit(fit, cont_mat)
  fit <- eBayes(fit, trend = TRUE, robust = TRUE)
  print(plotSA(fit))

  tests <- decideTests(fit, method = "global")
  write.fit(
    fit, tests, file.path(cache_dir, "results.tsv"),
    adjust = "BH", F.adjust = "BH", method = "global"
  )
  deg_list[[i]] <- read.delim(file.path(cache_dir, "results.tsv"))
}

results <- rbind(gse_list[[1]], gse_list[[2]], gse_list[[3]], gse_list[[4]])
gse_corrections <- data.frame(mean = results[ , 2], p = results[ , 5])
gse_corrections <- ihw(p ~ mean,  data = gse_corrections, alpha = 0.05)
gse_corrections <- adj_pvalues(gse_corrections)

results <- rbind(deg_list[[1]], deg_list[[2]], deg_list[[3]], deg_list[[4]])
deg_corrections <- data.frame(mean = results[ , 2], p = results[ , 5])
deg_corrections <- ihw(p ~ mean,  data = deg_corrections, alpha = 0.05)
deg_corrections <- adj_pvalues(deg_corrections)

rds <- file.path(cache_dir, "gse_data.rds")
if (!file.exists(rds)) {
  saveRDS(gse_data, file = rds)
}
rds <- file.path(cache_dir, "deg_data.rds")
if (!file.exists(rds)) {
  saveRDS(deg_data, file = rds)
}
rds <- file.path(cache_dir, "gse_list.rds")
if (!file.exists(rds)) {
  saveRDS(gse_list, file = rds)
}
rds <- file.path(cache_dir, "deg_list.rds")
if (!file.exists(rds)) {
  saveRDS(deg_list, file = rds)
}
rds <- file.path(cache_dir, "gse_corrections.rds")
if (!file.exists(rds)) {
  saveRDS(gse_corrections, file = rds)
}
rds <- file.path(cache_dir, "deg_corrections.rds")
if (!file.exists(rds)) {
  saveRDS(deg_corrections, file = rds)
}
```

# References

This is the concluding section of the document, where we output the `sessionInfo`, and create a bibliography for works cited.

```{r}
sessionInfo()
```
