---
title: "Otero-Garcia 2022 ssRNAseq - 03 Testing"
author:
  - name: "Emir Turkes"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
link-citations: true
output:
  html_document:
    code_folding: hide
    number_sections: true
    theme: lumen
    highlight: haddock
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
knit: (function(inputFile, encoding) {
  rmarkdown::render(
    inputFile, encoding = encoding,
    output_file = file.path(
      "..", "..", "results",
      "comb", "otero_garcia_2022_ssRNAseq_03testing.html"
    ))})
---

<style type="text/css">
body {font-size: 16px;}
h1.title {font-size: 35px;}
h1 {font-size: 24px;}
h2 {font-size: 22px;}
.toc-content {padding-left: 0px; padding-right: 0px;}
div.tocify {width: 100%;}
.tocify-subheader .tocify-item {
  font-size: 0.95em; padding-left: 25px; text-indent: 0;
}
.tocify-subheader .tocify-subheader .tocify-item {
  font-size: 0.95em; padding-left: 35px; text-indent: 0;
}
div.main-container {max-width: none; width: 100%;}
</style>

*This file is a part of [otero-garcia-2022-ssRNAseq](https://github.com/eturkes/otero-garcia-2022-ssRNAseq).*

The table of contents in the top left is clickable and can be used to quickly navigate the document.
To toggle the visibility of code, use the `CODE` toggles at the top right of chunks.
The toggle at the start of the document controls the visibility of all chunks.

```{r}
#    This file is part of otero-garcia-2022-ssRNAseq.
#    Copyright (C) 2024-2025  Emir Turkes, UK DRI at UCL
#
#    This program is free software: you can redistribute it and/or modify
#    it under the terms of the GNU General Public License as published by
#    the Free Software Foundation, either version 3 of the License, or
#    (at your option) any later version.
#
#    This program is distributed in the hope that it will be useful,
#    but WITHOUT ANY WARRANTY; without even the implied warranty of
#    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
#    GNU General Public License for more details.
#
#    You should have received a copy of the GNU General Public License
#    along with this program.  If not, see <http://www.gnu.org/licenses/>.
#
#    Emir Turkes can be contacted at emir.turkes@eturkes.com

# Load required packages, suppressing startup messages.
# -----------------------------------------------------
library(conflicted)
conflicts_prefer(edgeR::cpm, .quiet = TRUE)
packages <- c("GSEABase", "Seurat", "scuttle", "edgeR", "IHW")
invisible(
  suppressPackageStartupMessages(
    lapply(packages, FUN = library, character.only = TRUE)
  )
)
# -----------------------------------------------------

# Define global settings.
# -----------------------
knitr::opts_chunk$set(fig.width = 10, fig.height = 7, dpi = 96)
# -----------------------

# Useful variables.
# -----------------
name <- "comb"
step <- 3

cache_dir <- file.path("..", "..", "cache", name, paste0("step", "0", step))
if (!dir.exists(cache_dir)) {
  dir.create(cache_dir, recursive = TRUE)
}
# -----------------
```

# Prep

```{r}
seurat <- readRDS(file.path(cache_dir, "..", "step02", "final_seurat.rds"))
seurat

ctypes <- list("EX L2-4 CUX2", "EX L4-5 RORB", "EX L5 RORB", "EX L5-6 THEMIS")
seurat <- subset(seurat, idents = ctypes)
gc()
seurat

gene_sets <- readRDS(
  file.path(cache_dir, "..", "step02", "gene_sets.rds")
)
gene_sets_keep <- readRDS(
  file.path(cache_dir, "..", "step02", "gene_sets_keep.rds")
)
gene_anno <- readRDS(file.path(cache_dir, "..", "step02", "gene_anno.rds"))
```

# Testing

```{r}
gse <- as.SingleCellExperiment(seurat, assay = "GF")
# counts(gse) <- pmax(counts(gse), 0)
gse <- gse[rownames(gse) %in% names(gene_sets_keep), ]
gse <- suppressWarnings(
  aggregateAcrossCells(
    gse, colData(gse)[ , "donor_tau"],
    use_exprs_values = "counts", statistics = "sum"
  )
)
gse <- gse[ , order(gse$tau)]

deg <- as.SingleCellExperiment(seurat, assay = "RNA")
deg <- suppressWarnings(
  aggregateAcrossCells(
    deg, colData(deg)[ , "donor_tau"],
    use_exprs_values = "counts", statistics = "sum"
  )
)
deg <- deg[ , order(deg$tau)]

dge_orig <- DGEList(counts(gse))
keep <- filterByExpr(
  dge_orig, group = gse$tau,
  min.count = 10, min.total.count = 15
)
dge_orig <- dge_orig[keep, ]
dge_orig <- calcNormFactors(dge_orig)
dge <- DGEList(counts(gse))
dge$samples$norm.factors <- dge_orig$samples$norm.factors
keep <- filterByExpr(
  dge, group = gse$tau,
  min.count = 10, min.total.count = 15
)
dge <- dge[keep, , keep.lib.sizes = FALSE]
dge <- calcNormFactors(dge)

gse <- gse[rownames(gse) %in% rownames(dge)]
logcounts(gse, withDimnames = FALSE) <- cpm(
  dge, log = TRUE, prior.count = 6
)

design <- model.matrix(~ 0 + gse$tau)
colnames(design) <- make.names(colnames(design))
cont_mat <- makeContrasts(
  NFT_CTRL =
    gse.tauNFT -
    gse.tauCTRL,
  levels = design
)

corr <- duplicateCorrelation(logcounts(gse), design, block = gse$donor)
fit <- lmFit(
  logcounts(gse), design,
  correlation = corr$consensus.correlation, block = gse$donor
)
fit <- contrasts.fit(fit, cont_mat)
fit <- eBayes(fit, trend = TRUE, robust = TRUE)
plotSA(fit)

tests <- decideTests(fit)
tc <- textConnection("results", open = "w")
write.fit(fit, tests, file = tc, adjust = "BH")
close(tc)
results_gse <- read.delim(text = results)

rownames(results_gse) <- results_gse$X
results_gse <- results_gse[ , 2:6]
results_gse$GeneSet <- rownames(results_gse)
results_gse <- results_gse[order(results_gse$P.value), ]

gene_sets_sub <- gene_sets[names(gene_sets) %in% rownames(gse)]
gene_sets_sub <- unique(unlist(geneIds(gene_sets_sub)))
deg <- deg[rownames(deg) %in% gene_sets_sub, ]
dge_orig <- DGEList(counts(deg))
keep <- filterByExpr(
  dge_orig, group = deg$tau,
  min.count = 10, min.total.count = 15
)
dge_orig <- dge_orig[keep, ]
dge_orig <- calcNormFactors(dge_orig)
dge <- DGEList(counts(deg))
dge$samples$norm.factors <- dge_orig$samples$norm.factors
keep <- filterByExpr(
  dge, group = deg$tau,
  min.count = 10, min.total.count = 15
)
dge <- dge[keep, , keep.lib.sizes = FALSE]
dge <- calcNormFactors(dge)

deg <- deg[rownames(deg) %in% rownames(dge)]
logcounts(deg, withDimnames = FALSE) <- cpm(
  dge, log = TRUE, prior.count = 4
)

corr <- duplicateCorrelation(logcounts(deg), design, block = deg$donor)
fit <- lmFit(
  logcounts(deg), design,
  correlation = corr$consensus.correlation, block = deg$donor
)
fit <- contrasts.fit(fit, cont_mat)
fit <- eBayes(fit, trend = TRUE, robust = TRUE)
plotSA(fit)

tests <- decideTests(fit)
tc <- textConnection("results", open = "w")
write.fit(fit, tests, file = tc, adjust = "BH")
close(tc)
results_deg <- read.delim(text = results)

rownames(results_deg) <- results_deg$X
results_deg <- results_deg[ , 2:6]
gene_anno_sub <- gene_anno[gene_anno$ensembl %in% rownames(results_deg), ]
gene_anno_sub <- gene_anno_sub[
  match(rownames(results_deg), gene_anno_sub$ensembl),
]
results_deg$Gene <- gene_anno_sub$symbol
results_deg <- results_deg[ , c(ncol(results_deg), 1:(ncol(results_deg) - 1))]
results_deg <- results_deg[order(results_deg$P.value), ]

meta <- data.frame(donor = gse$donor, tau = gse$tau)
gse <- removeBatchEffect(logcounts(gse), batch = gse$donor, group = gse$tau)
deg <- removeBatchEffect(logcounts(deg), batch = deg$donor, group = deg$tau)

rds <- file.path(cache_dir, "meta.rds")
if (!file.exists(rds)) {
  saveRDS(meta, file = rds)
}
rds <- file.path(cache_dir, "gse.rds")
if (!file.exists(rds)) {
  saveRDS(gse, file = rds)
}
rds <- file.path(cache_dir, "deg.rds")
if (!file.exists(rds)) {
  saveRDS(deg, file = rds)
}
rds <- file.path(cache_dir, "res_gse.rds")
if (!file.exists(rds)) {
  saveRDS(results_gse, file = rds)
}
rds <- file.path(cache_dir, "res_deg.rds")
if (!file.exists(rds)) {
  saveRDS(results_deg, file = rds)
}
```

# References

This is the concluding section of the document, where we output the `sessionInfo`, and create a bibliography for works cited.

```{r}
sessionInfo()
```
