---
title: "Otero-Garcia 2022 ssRNAseq - 01 Prep"
author:
  - name: "Emir Turkes"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
link-citations: true
output:
  html_document:
    code_folding: hide
    number_sections: true
    theme: lumen
    highlight: haddock
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
knit: (function(inputFile, encoding) {
  rmarkdown::render(
    inputFile, encoding = encoding,
    output_file = file.path(
      "..", "..", "results",
      "comb", "otero_garcia_2022_ssRNAseq_01prep.html"
    ))})
---

<style type="text/css">
body {font-size: 16px;}
h1.title {font-size: 35px;}
h1 {font-size: 24px;}
h2 {font-size: 22px;}
.toc-content {padding-left: 0px; padding-right: 0px;}
div.tocify {width: 100%;}
.tocify-subheader .tocify-item {
  font-size: 0.95em; padding-left: 25px; text-indent: 0;
}
.tocify-subheader .tocify-subheader .tocify-item {
  font-size: 0.95em; padding-left: 35px; text-indent: 0;
}
div.main-container {max-width: none; width: 100%;}
</style>

*This file is a part of [otero-garcia-2022-ssRNAseq](https://github.com/eturkes/otero-garcia-2022-ssRNAseq).*

The table of contents in the top left is clickable and can be used to quickly navigate the document.
To toggle the visibility of code, use the `CODE` toggles at the top right of chunks.
The toggle at the start of the document controls the visibility of all chunks.

```{r}
#    This file is part of otero-garcia-2022-ssRNAseq.
#    Copyright (C) 2024  Emir Turkes, UK DRI at UCL
#
#    This program is free software: you can redistribute it and/or modify
#    it under the terms of the GNU General Public License as published by
#    the Free Software Foundation, either version 3 of the License, or
#    (at your option) any later version.
#
#    This program is distributed in the hope that it will be useful,
#    but WITHOUT ANY WARRANTY; without even the implied warranty of
#    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
#    GNU General Public License for more details.
#
#    You should have received a copy of the GNU General Public License
#    along with this program.  If not, see <http://www.gnu.org/licenses/>.
#
#    Emir Turkes can be contacted at emir.turkes@eturkes.com

# Load required packages, suppressing startup messages.
# -----------------------------------------------------
library(conflicted)
packages <- c(
  "Seurat", "parallelly", "ggplot2", "dplyr", "ggrepel", "viridis"
)
invisible(
  suppressPackageStartupMessages(
    lapply(packages, FUN = library, character.only = TRUE)
  )
)
# -----------------------------------------------------

# Define global settings.
# -----------------------
knitr::opts_chunk$set(fig.width = 10, fig.height = 7, dpi = 96)
# -----------------------

# Define functions.
# -----------------
source(file.path("..", "utils.R"))
# -----------------

# Useful variables.
# -----------------
name <- "comb"
step <- 1

data_dir <- file.path("..", "..", "data")
cache_dir <- file.path("..", "..", "cache", name, paste0("step", "0", step))
if (!dir.exists(cache_dir)) {
  dir.create(cache_dir, recursive = TRUE)
}
# -----------------

# Metadata to plot after dimensionality reduction and clustering.
# Values in list can include "no_legend and/or "no_label" to exclude those.
# -------------------------------------------------------------------------
metadata_to_plot <- vector("list", length = 3)
names(metadata_to_plot) <- c("seurat_clusters", "donor", "tau")
metadata_to_plot[[2]] <- "no_label"
metadata_to_plot[[3]] <- "no_label"
# -------------------------------------------------------------------------
```

# Merging

Merge individual samples into a single Seurat object.

```{r}
rds <- file.path(cache_dir, "merged_seurat.rds")
if (file.exists(rds)) {
  seurat <- readRDS(rds)
} else {

  file_list <- list.files(file.path(cache_dir, "..", ".."), recursive = TRUE)
  file_list <- file_list[grep("seurat_prep.rds", file_list)]
  files <- vector("list", length(file_list))
  for (i in seq_along(files)) {
    files[[i]] <- readRDS(file.path(cache_dir, "..", "..", file_list[i]))
    files[[i]]@meta.data[grep("SCT", colnames(files[[i]]@meta.data))] <- NULL
  }

  gc()
  seurat <- merge(
    files[[2]], c(files[[4]], files[[6]], files[[8]],
    files[[10]], files[[12]], files[[14]], files[[16]],
    files[[1]], files[[3]], files[[5]], files[[7]],
    files[[9]], files[[11]], files[[13]], files[[15]]
    )
  )
  gc()

  DefaultAssay(seurat) <- "RNA"
  seurat <- JoinLayers(seurat)

  gc()
  plan("sequential")
  seurat <- PrepSCTFindMarkers(seurat, verbose = FALSE)
  gc()

  seurat[["PREP_SCT"]] <- seurat[["SCT"]]
  seurat[["SCT"]] <- NULL

  seurat@misc <- files[[1]]@misc

  seurat$donor <- factor(
    c(
      rep("Donor1", times = ncol(files[[2]])),
      rep("Donor2", times = ncol(files[[4]])),
      rep("Donor3", times = ncol(files[[6]])),
      rep("Donor4", times = ncol(files[[8]])),
      rep("Donor5", times = ncol(files[[10]])),
      rep("Donor6", times = ncol(files[[12]])),
      rep("Donor7", times = ncol(files[[14]])),
      rep("Donor8", times = ncol(files[[16]])),
      rep("Donor1", times = ncol(files[[1]])),
      rep("Donor2", times = ncol(files[[3]])),
      rep("Donor3", times = ncol(files[[5]])),
      rep("Donor4", times = ncol(files[[7]])),
      rep("Donor5", times = ncol(files[[9]])),
      rep("Donor6", times = ncol(files[[11]])),
      rep("Donor7", times = ncol(files[[13]])),
      rep("Donor8", times = ncol(files[[15]]))
    )
  )
  seurat$tau <- factor(
    c(
      rep("CTRL", times = ncol(files[[2]])),
      rep("CTRL", times = ncol(files[[4]])),
      rep("CTRL", times = ncol(files[[6]])),
      rep("CTRL", times = ncol(files[[8]])),
      rep("CTRL", times = ncol(files[[10]])),
      rep("CTRL", times = ncol(files[[12]])),
      rep("CTRL", times = ncol(files[[14]])),
      rep("CTRL", times = ncol(files[[16]])),
      rep("NFT", times = ncol(files[[1]])),
      rep("NFT", times = ncol(files[[3]])),
      rep("NFT", times = ncol(files[[5]])),
      rep("NFT", times = ncol(files[[7]])),
      rep("NFT", times = ncol(files[[9]])),
      rep("NFT", times = ncol(files[[11]])),
      rep("NFT", times = ncol(files[[13]])),
      rep("NFT", times = ncol(files[[15]]))
    )
  )

  rm(files)
  saveRDS(seurat, file = rds)
}

seurat
```

# Cluster Pipeline

Perform dimensionality reduction and clustering using the intersect of SCTransform scaled data between samples.

```{r}
rds <- file.path(cache_dir, "clustered_seurat.rds")
if (file.exists(rds)) {
  seurat <- readRDS(rds)
} else {

  gc()
  seurat <- SCTransform(seurat, vars.to.regress = "donor", verbose = FALSE)
  gc()

  seurat <- RunPCA(seurat, verbose = FALSE)
  add_df <- data.frame(Embeddings(seurat)[ , 1:2])
  names(add_df) <- paste0("pca", seq(ncol(add_df)))
  seurat$pca1 <- add_df$pca1
  seurat$pca2 <- add_df$pca2
  reduction <- "pca"
  dims <- 1:33

  seurat <- RunUMAP(seurat, dims = dims, reduction = reduction, verbose = FALSE)
  add_df <- data.frame(Embeddings(seurat, reduction = "umap"))
  names(add_df) <- paste0("umap", seq(ncol(add_df)))
  seurat$umap1 <- add_df$umap1
  seurat$umap2 <- add_df$umap2

  seurat <- FindNeighbors(
    seurat, reduction = reduction, dims = dims, verbose = FALSE
  )
  seurat <- FindClusters(seurat, resolution = 0.005, verbose = FALSE)

  saveRDS(seurat, file = rds)
}

seurat

ElbowPlot(seurat, ndims = 50) + geom_vline(xintercept = 33)

for (i in seq_along(metadata_to_plot)) {
  if (
    "no_legend" %in% metadata_to_plot[[i]] &&
      "no_label" %in% metadata_to_plot[[i]]
  ) {
    print(
      red_dim_plot(
        seurat, x = "umap1", y = "umap2", color = names(metadata_to_plot)[i]
      ) +
        NoLegend() +
        ggtitle(names(metadata_to_plot)[i])
    )
  } else if ("no_legend" %in% metadata_to_plot[[i]]) {
    print(
      red_dim_plot(
        seurat, x = "umap1", y = "umap2",
        color = names(metadata_to_plot)[i], type = "cat"
      ) +
        NoLegend() +
        ggtitle(names(metadata_to_plot)[i])
    )
  } else if ("no_label" %in% metadata_to_plot[[i]]) {
    print(
      red_dim_plot(
        seurat, x = "umap1", y = "umap2", color = names(metadata_to_plot)[i]
      ) +
        ggtitle(names(metadata_to_plot)[i])
    )
  } else {
    print(
      red_dim_plot(
        seurat, x = "umap1", y = "umap2",
        color = names(metadata_to_plot)[i], type = "cat"
      ) +
        ggtitle(names(metadata_to_plot)[i])
    )
  }
}

red_dim_plot(
  seurat, x = "umap1", y = "umap2", color = "nFeature_SCT", type = "cont"
) + ggtitle("nFeature_SCT")
red_dim_plot(
  seurat, x = "umap1", y = "umap2", color = "nCount_SCT", type = "cont"
) + ggtitle("nCount_SCT")
red_dim_plot(
  seurat, x = "umap1", y = "umap2", color = "nFeature_RNA", type = "cont"
) + ggtitle("nFeature_RNA")
red_dim_plot(
  seurat, x = "umap1", y = "umap2", color = "nCount_RNA", type = "cont"
) + ggtitle("nCount_RNA")

red_dim_plot(
  seurat, x = "umap1", y = "umap2", color = "percent_mt", type = "cont"
) + ggtitle("percent_mt")
red_dim_plot(
  seurat, x = "umap1", y = "umap2", color = "percent_ribo", type = "cont"
) + ggtitle("percent_ribo")
red_dim_plot(
  seurat, x = "umap1", y = "umap2", color = "percent_ribo", type = "cont"
) + ggtitle("percent_malat1")

VlnPlot(
  seurat,
  features = c("nFeature_RNA", "nCount_RNA"),
  ncol = 2
)
VlnPlot(
  seurat,
  features = c("nFeature_SCT", "nCount_SCT"),
  ncol = 2
)
VlnPlot(
  seurat,
  features = c("percent_mt", "percent_ribo"),
  ncol = 2
)
```

# Data Alignment

Remove GABAergic cells then use the author's Seurat object as a reference to map their cell-types onto ours.

```{r}
FeaturePlot(
  seurat, "ENSG00000104888", order = TRUE, cols = c("lightgrey", "red")
) + ggtitle("SLC17A7")
FeaturePlot(
  seurat, "ENSG00000128683", order = TRUE, cols = c("lightgrey", "red")
) + ggtitle("GAD1")
FeaturePlot(
  seurat, "ENSG00000131095", order = TRUE, cols = c("lightgrey", "red")
) + ggtitle("GFAP")
FeaturePlot(
  seurat, "ENSG00000168314", order = TRUE, cols = c("lightgrey", "red")
) + ggtitle("MOBP")
FeaturePlot(
  seurat, "ENSG00000184221", order = TRUE, cols = c("lightgrey", "red")
) + ggtitle("OLIG1")
FeaturePlot(
  seurat, "ENSG00000100362", order = TRUE, cols = c("lightgrey", "red")
) + ggtitle("PVALB")
FeaturePlot(
  seurat, "ENSG00000172137", order = TRUE, cols = c("lightgrey", "red")
) + ggtitle("CALB2")
FeaturePlot(
  seurat, "ENSG00000125869", order = TRUE, cols = c("lightgrey", "red")
) + ggtitle("LAMP5")
FeaturePlot(
  seurat, "ENSG00000157005", order = TRUE, cols = c("lightgrey", "red")
) + ggtitle("SST")

rds <- file.path(cache_dir, "annotated_seurat.rds")
rds2 <- file.path(cache_dir, "seurat_sub.rds")
if (file.exists(rds) & file.exists(rds2)) {
  seurat <- readRDS(rds)
  seurat_sub <- readRDS(rds2)
} else {

  seurat_sub <- subset(seurat, idents = 0)

  ref <- readRDS(file.path(data_dir, "excitatory.rds"))
  ref <- ref[ , -grep("CTRL", x = ref$donor_id)]
  ref$donor_id <- factor(ref$donor_id)

  gc()
  ref <- SCTransform(ref, vars.to.regress = "donor_id", verbose = FALSE)
  gc()
  ref <- RunPCA(ref, verbose = FALSE)

  anchors <- FindTransferAnchors(
    ref, seurat_sub, normalization.method = "SCT",
    recompute.residuals = FALSE, reference.reduction = "pca", dims = 1:33
  )
  predictions <- MapQuery(
    anchors, seurat_sub, ref,
    refdata = "Cell.Types", reference.reduction = "umap"
  )
  seurat_sub$predicted_ctype <- factor(predictions$predicted.id)
  rm(ref, predictions, anchors)
  gc()

  seurat <- FindClusters(seurat, resolution = 0.2, verbose = FALSE)
  
  seurat <- RenameIdents(
    object = seurat, "0" = "EX L2-4 CUX2", "1" = "EX L5 RORB",
    "2" = "EX L4-5 RORB", "3" = "EX L5-6 THEMIS", "4" = "IN PVALB",
    "5" = "IN CALB2", "6" = "EX L6 FEZF2", "7" = "IN LAMP5",
    "8" = "EX L5 FEZF2", "9" = "Astrocyte", "10" = "EX L6 THEMIS",
    "11" = "Oligodendrocyte", "12" = "OPC", "13" = "IN SST"
  )
  seurat@active.ident <- factor(as.character(seurat@active.ident))
  seurat$ctype <- seurat@active.ident

  saveRDS(seurat, file = rds)
  saveRDS(seurat_sub, file = rds2)
}

red_dim_plot(
  seurat_sub, x = "umap1", y = "umap2", color = "predicted_ctype", type = "cat"
)
red_dim_plot(
  seurat, x = "umap1", y = "umap2", color = "ctype", type = "cat"
)
```

# References

This is the concluding section of the document, where we output the `sessionInfo`, and create a bibliography for works cited.

```{r}
sessionInfo()
```
