---
title: "Otero-Garcia 2022 ssRNAseq - 02 GeneFunnel"
author:
  - name: "Emir Turkes"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
link-citations: true
output:
  html_document:
    code_folding: hide
    number_sections: true
    theme: lumen
    highlight: haddock
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
knit: (function(inputFile, encoding) {
  rmarkdown::render(
    inputFile, encoding = encoding,
    output_file = file.path(
      "..", "results", "otero_garcia_2022_ssRNAseq_02genefunnel.html"
    ))})
---

<style type="text/css">
body {font-size: 16px;}
h1.title {font-size: 35px;}
h1 {font-size: 24px;}
h2 {font-size: 22px;}
.toc-content {padding-left: 0px; padding-right: 0px;}
div.tocify {width: 100%;}
.tocify-subheader .tocify-item {
  font-size: 0.95em; padding-left: 25px; text-indent: 0;
}
.tocify-subheader .tocify-subheader .tocify-item {
  font-size: 0.95em; padding-left: 35px; text-indent: 0;
}
div.main-container {max-width: none; width: 100%;}
</style>

*This file is a part of [otero-garcia-2022-ssRNAseq](https://github.com/eturkes/otero-garcia-2022-ssRNAseq).*

The table of contents in the top left is clickable and can be used to quickly navigate the document.
To toggle the visibility of code, use the `CODE` toggles at the top right of chunks.
The toggle at the start of the document controls the visibility of all chunks.

```{r}
#    This file is part of otero-garcia-2022-ssRNAseq.
#    Copyright (C) 2024  Emir Turkes, UK DRI at UCL
#
#    This program is free software: you can redistribute it and/or modify
#    it under the terms of the GNU General Public License as published by
#    the Free Software Foundation, either version 3 of the License, or
#    (at your option) any later version.
#
#    This program is distributed in the hope that it will be useful,
#    but WITHOUT ANY WARRANTY; without even the implied warranty of
#    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
#    GNU General Public License for more details.
#
#    You should have received a copy of the GNU General Public License
#    along with this program.  If not, see <http://www.gnu.org/licenses/>.
#
#    Emir Turkes can be contacted at emir.turkes@eturkes.com

# Load required packages, suppressing startup messages.
# -----------------------------------------------------
library(conflicted)
conflicts_prefer(edgeR::cpm, .quiet = TRUE)
packages <- c(
  "GSEABase", "Seurat", "scuttle", "edgeR", "DT", "GSVA", "Rcpp", "biomaRt",
  "IHW"
)
invisible(
  suppressPackageStartupMessages(
    lapply(packages, FUN = library, character.only = TRUE)
  )
)
# -----------------------------------------------------

# Define global settings.
# -----------------------
knitr::opts_chunk$set(fig.width = 10, fig.height = 7, dpi = 300)
# -----------------------

# Define functions.
# -----------------
source("utils.R")
sourceCpp(file.path("..", "src", "calculateScores.cpp"))

`%notin%` <- Negate(`%in%`)
# -----------------

# Useful variables.
# -----------------
step <- 2

cache_dir <- file.path("..", "cache", paste0("step", "0", step))
if (!dir.exists(cache_dir)) {
  dir.create(cache_dir, recursive = TRUE)
}
# -----------------
```

# Prep

```{r}
seurat_orig <- readRDS(file.path(cache_dir, "..", "step01", "seurat_prep.rds"))

rds <- file.path(cache_dir, "gene_sets.rds")
if (file.exists(rds)) {
  gene_sets <- readRDS(rds)
} else {

  gene_sets <- getGmt(
    file.path(
      "..", "data",
      "gprofiler_hsapiens_ENSG_2024_08_25", "hsapiens.GO.comb.ENSG.gmt"
    )
  )
  keep <- filterGeneSets(gene_sets, min.sz = 2)
  gene_sets <- gene_sets[names(gene_sets) %in% names(keep)]

  remove <- unique(unlist(geneIds(gene_sets)))[
    unique(unlist(geneIds(gene_sets))) %notin% rownames(seurat_orig)
  ]
  remove <- which(
    sapply(geneIds(gene_sets), FUN = function(x) any(x %in% remove))
  )
  gene_sets <- gene_sets[-remove]

  for (i in seq_along(gene_sets@.Data)) {
    go_id <- gene_sets[[i]]@setName
    suppressWarnings(gene_sets[[i]]@setName <- gene_sets[[i]]@shortDescription)
    suppressWarnings(gene_sets[[i]]@shortDescription <- go_id)
  }

  gene_sets_keep <- computeGeneSetsOverlapMax(
    gene_sets, uniqGenes = unique(unlist(geneIds(gene_sets)))
  )
  gene_sets_keep <- rowSums(gene_sets_keep)
  gene_sets_keep <- gene_sets_keep[order(gene_sets_keep, decreasing = TRUE)]
  gene_sets_keep <- gene_sets[match(names(gene_sets_keep), names(gene_sets))]

  dup <- sapply(
    geneIds(gene_sets_keep), FUN = function(x) paste(sort(x), collapse = "-")
  )
  dup <- which(duplicated(dup))
  gene_sets <- gene_sets_keep[-dup]

  saveRDS(gene_sets, file = rds)
}
gene_sets

rds <- file.path(cache_dir, "gene_sets_keep.rds")
if (file.exists(rds)) {
  gene_sets_keep <- readRDS(rds)
} else {

  keep <- filterGeneSets(gene_sets, min.sz = 2, max.sz = 45)
  gene_sets_keep <- gene_sets[names(gene_sets) %in% names(keep)]

  gene_sets_keep <- gene_sets_keep[
    !grepl("regulation", x = names(gene_sets_keep)) |
      grepl("positive regulation", x = names(gene_sets_keep))
  ]

  gene_sets_keep <- gene_sets_keep[
    !grepl("selection", x = names(gene_sets_keep)) |
      grepl("positive selection", x = names(gene_sets_keep))
  ]

  mart <- useEnsembl("ensembl", "hsapiens_gene_ensembl")
  attributes <- c("ensembl_gene_id", "chromosome_name")
  gene_anno <- getBM(
    attributes, filters = "ensembl_gene_id",
    values = unique(unlist(geneIds(gene_sets_keep))), mart = mart
  )
  remove <- gene_anno[
    gene_anno$chromosome_name == "Y" |
      gene_anno$ensembl_gene_id == "ENSG00000229807" |
      gene_anno$ensembl_gene_id == "ENSG00000270641",
  ]$ensembl_gene_id
  remove <- which(
    sapply(geneIds(gene_sets_keep), FUN = function(x) any(x %in% remove))
  )
  gene_sets_keep <- gene_sets_keep[-remove]

  gene_sets_ordered <- computeGeneSetsOverlapMax(
    gene_sets_keep, uniqGenes = unique(unlist(geneIds(gene_sets_keep)))
  )
  gene_sets_ordered <- rowSums(gene_sets_ordered)
  gene_sets_ordered <- gene_sets_ordered[order(gene_sets_ordered)]
  gene_sets_keep <- gene_sets_keep[
    match(names(gene_sets_ordered), names(gene_sets_keep))
  ]

  gene_sets_keep <- computeGeneSetsOverlapMax(
    gene_sets_keep, uniqGenes = unique(unlist(geneIds(gene_sets_keep)))
  )
  gene_sets_keep[upper.tri(gene_sets_keep)] <- 0
  diag(gene_sets_keep) <- 0
  gene_sets_keep <- apply(gene_sets_keep, MARGIN = 1, FUN = max)
  gene_sets_keep <- gene_sets_keep[gene_sets_keep < 0.7]

  rm(gene_sets_ordered)
  saveRDS(gene_sets_keep, file = rds)
}
gene_sets[names(gene_sets) %in% names(gene_sets_keep)]

rds <- file.path(cache_dir, "gene_anno.rds")
if (!file.exists(rds)) {

  mart <- useEnsembl("ensembl", "hsapiens_gene_ensembl")
  attributes <- c("ensembl_gene_id", "external_gene_name")
  gene_anno <- getBM(
    attributes, filters = "ensembl_gene_id",
    values = rownames(seurat_orig), mart = mart
  )

  add <- data.frame(
    ensembl_gene_id = rownames(seurat_orig)[
      rownames(seurat_orig) %notin% gene_anno$ensembl_gene_id
    ]
  )
  add$external_gene_name <- add$ensembl_gene_id
  gene_anno <- rbind(gene_anno, add)

  replace <- gene_anno[gene_anno$external_gene_name == "", ]
  for (i in seq(nrow(replace))) {
    for (j in seq(nrow(gene_anno))) {
      if (replace$ensembl_gene_id[i] == gene_anno$ensembl_gene_id[j]) {
        gene_anno$external_gene_name[j] <- gene_anno$ensembl_gene_id[j]
      }
    }
  }

  replace <- gene_anno[
    duplicated(gene_anno$external_gene_name) |
      duplicated(gene_anno$external_gene_name, fromLast = TRUE),
  ]
  for (i in seq(nrow(replace))) {
    for (j in seq(nrow(gene_anno))) {
      if (replace$ensembl_gene_id[i] == gene_anno$ensembl_gene_id[j]) {
        gene_anno$external_gene_name[j] <-
          paste(
            gene_anno$external_gene_name[j], gene_anno$ensembl_gene_id[j],
            sep = "-"
          )
      }
    }
  }

  gene_anno <- gene_anno[ , c(2, 1)]
  colnames(gene_anno) <- c("symbol", "ensembl")
  saveRDS(gene_anno, file = rds)
}

ctypes <- vector("list", length = 1)
ctypes <- list(
  `L2/4_Cortex` = c("Ex01_CUX2-LAMP5 (L2-L3)", "Ex02_CUX2-COL5A2 (L2-L4)")
)

rds <- file.path(cache_dir, "seurat_list.rds")
if (file.exists(rds)) {
  rm(seurat_orig)
  gc()
  seurat_list <- readRDS(rds)
} else {

  seurat_orig@active.ident <- seurat_orig$Cell.Types
  seurat_list <- vector("list", length = 1)
  for (i in seq_along(ctypes)) {

    seurat <- subset(seurat_orig, idents = ctypes[[i]])
    seurat$tau <- ifelse(seurat$SORT == "AT8", yes = "NFT", no = "CTRL")
    seurat$Sample.ID <- factor(seurat$Sample.ID)
    seurat$donor_id <- factor(seurat$donor_id)

    gc()
    DefaultAssay(seurat) <- "RNA"
    scores <- calculateScores(
      GetAssayData(seurat, layer = "counts"), rownames(seurat),
      geneIds(gene_sets)
    )
    gc()
    rownames(scores) <- names(gene_sets)
    colnames(scores) <- colnames(seurat)

    seurat[["GF"]] <- CreateAssay5Object(scores, data = log1p(scores))
    DefaultAssay(seurat) <- "GF"
    seurat$nCount_GF <- colSums(seurat, layer = "counts")
    seurat$nFeature_GF <- colSums(
      GetAssayData(seurat, layer = "counts") > 0
    )

    seurat_list[[i]] <- seurat
  }
  rm(seurat_orig, seurat, scores)
  gc()
  saveRDS(seurat_list, file = rds)
}
seurat_list

gse_data <- vector("list", length = 1)
deg_data <- vector("list", length = 1)
gse_list <- vector("list", length = 1)
deg_list <- vector("list", length = 1)

for (i in seq_along(seurat_list)) {

  seurat_sub <- subset(
    seurat_list[[i]], downsample = min(table(seurat_list[[i]]$Sample.ID))
  )

  gse <- as.SingleCellExperiment(seurat_list[[i]], assay = "GF")
  keep <- filterByExpr(
    GetAssayData(seurat_sub, assay = "GF", layer = "counts"),
    group = seurat_sub$tau,
    min.prop = 0.01,
    min.count = 1,
    min.total.count = 1.5,
    large.n = 0
  )
  gse <- gse[keep, ]
  gse <- suppressWarnings(
    aggregateAcrossCells(
      gse, colData(gse)[ , "Sample.ID"],
      use_exprs_values = "counts", statistics = "sum"
    )
  )
  gse <- gse[ , order(gse$tau)]

  deg <- as.SingleCellExperiment(seurat_list[[i]], assay = "RNA")
  keep <- filterByExpr(
    GetAssayData(seurat_sub, assay = "RNA", layer = "counts"),
    group = seurat_sub$tau,
    min.prop = 0.01,
    min.count = 1,
    min.total.count = 1.5,
    large.n = 0
  )
  deg <- deg[keep, ]
  deg <- suppressWarnings(
    aggregateAcrossCells(
      deg, colData(deg)[ , "Sample.ID"],
      use_exprs_values = "counts", statistics = "sum"
    )
  )
  deg <- deg[ , order(deg$tau)]

  dge <- DGEList(counts(gse))
  keep <- filterByExpr(
    dge, group = gse$tau,
    min.count = 10, min.total.count = 15
  )
  dge <- dge[keep, ]
  dge <- calcNormFactors(dge)
  keep <- filterByExpr(
    dge, group = gse$tau,
    min.count = 10, min.total.count = 15
  )
  dge <- dge[keep, , keep.lib.sizes = FALSE]

  gse <- gse[rownames(gse) %in% rownames(dge)]
  logcounts(gse, withDimnames = FALSE) <- cpm(
    dge, log = TRUE, prior.count = 2
  )

  gse <- gse[rownames(gse) %in% names(gene_sets_keep), ]
  gse_data[[i]] <- gse

  dge <- DGEList(counts(deg))
  keep <- filterByExpr(
    dge, group = deg$tau,
    min.count = 10, min.total.count = 15
  )
  dge <- dge[keep, ]
  dge <- calcNormFactors(dge)
  keep <- filterByExpr(
    dge, group = gse$tau,
    min.count = 10, min.total.count = 15
  )
  dge <- dge[keep, , keep.lib.sizes = FALSE]

  deg <- deg[rownames(deg) %in% rownames(dge)]
  logcounts(deg, withDimnames = FALSE) <- cpm(
    dge, log = TRUE, prior.count = 2
  )

  gene_sets_sub <- gene_sets[names(gene_sets) %in% rownames(gse)]
  deg <- deg[rownames(deg) %in% unique(unlist(geneIds(gene_sets_sub))), ]
  deg_data[[i]] <- deg

  design <- model.matrix(~ 0 + gse$tau + gse$donor_id)
  colnames(design) <- make.names(colnames(design))
  cont_mat <- makeContrasts(
    NFT_CTRL =
      gse.tauNFT -
      gse.tauCTRL,
    levels = design
  )

  fit <- lmFit(logcounts(gse), design)
  fit <- contrasts.fit(fit, cont_mat)
  fit <- eBayes(fit, trend = TRUE, robust = TRUE)

  tests <- decideTests(fit, method = "global")
  write.fit(
    fit, tests, file.path(cache_dir, "results.tsv"),
    adjust = "BH", F.adjust = "BH", method = "global"
  )
  gse_list[[i]] <- read.delim(file.path(cache_dir, "results.tsv"))

  fit <- lmFit(logcounts(deg), design)
  fit <- contrasts.fit(fit, cont_mat)
  fit <- eBayes(fit, trend = TRUE, robust = TRUE)

  tests <- decideTests(fit, method = "global")
  write.fit(
    fit, tests, file.path(cache_dir, "results.tsv"),
    adjust = "BH", F.adjust = "BH", method = "global"
  )
  deg_list[[i]] <- read.delim(file.path(cache_dir, "results.tsv"))
}

rds <- file.path(cache_dir, "gse_data.rds")
if (!file.exists(rds)) {
  saveRDS(gse_data, file = rds)
}
rds <- file.path(cache_dir, "deg_data.rds")
if (!file.exists(rds)) {
  saveRDS(deg_data, file = rds)
}
rds <- file.path(cache_dir, "gse_list.rds")
if (!file.exists(rds)) {
  saveRDS(gse_list, file = rds)
}
rds <- file.path(cache_dir, "deg_list.rds")
if (!file.exists(rds)) {
  saveRDS(deg_list, file = rds)
}
```

# References

This is the concluding section of the document, where we output the `sessionInfo`, and create a bibliography for works cited.

```{r}
sessionInfo()
```
